Testing the Relationship between Phenomenological Control related to Illusion Sensitivity

### 2024-06-04 paper/manuscript.qmd
Updated manuscript

Interestingly, @lush2022 did test for a relationship between PC and illusion sensitivity using the Müller-Lyer illusion (in which the arrangement of the arrowheads flanking two lines makes them appear as having different lengths), and reported evidence in favour of an absence of correlation between the two measures. This finding was interpreted as indicative of the cognitive impenetrability of illusions, implying that the effect is driven by low-level processes and therefore not influenced by top-down mechanisms such as PC. The goal of this study is thus to replicate the results from @lush2022 pointing to an absence of a relationship between phenomenological control and illusion sensitivity, by generalising them to a different illusion task that would encompass other illusion types (see @tbl-table1).

@@

The experiment's setup follows the born-open principle [@deleeuw2023]. The online experiment, implemented entirely using *JsPsych* [@de2015jspsych], has its code stored on GitHub and will leverage the power of the platform to host the experiment for free. Participants' raw data files are then automatically stored on a European server on OSF; and the deanonymized and ready-to-use data (downloaded from OSF and formatted using a reproducible script) will be made available in the same GitHub repository, alongside the scripts used for the analysis.

Participants will be presented with a consent form followed by demographic questions (gender, education level, age, and ethnicity). Participants will then be administered the PCS and the Illusion Game task (IG) in a counterbalanced order.

@@

The task is an adaptation of the one used in @makowski2023 to make it shorter and more reliable, in which participants must make perceptual judgments (e.g., "which red line is the longer") as quickly and accurately as possible. It includes 3 illusion types, namely Ebbinghaus, Müller-Lyer, and Vertical-Horizontal. The procedure encompasses 2 sets of 80 trials for each illusion type. Each set will include, in a random order, the 3 blocks of illusion types, in which trials are separated by a fixation cross, temporally (uniformly sampled duration of 500 - 1000s) and spatially jittered (around the centre of the screen in a radius of a 1 cm) to attenuate its potential usefulness as a reference point. After each illusion type block, a score is presented (computed as a scaled Inverse Efficiency Score) as a gamification mechanism to increase motivation to perform to the best of one's abilities.

@@

The two sets of 3 illusion blocks will be separated by 2 short questionnaires acting as a break, namely the IPIP-6 [@sibley2011mini], measuring 6 personality traits with 24 analogue scales items, and the PID-5 [@krueger2011], measuring 5 maladaptive personality traits with 25 Likert scales items.

>>>>

Interestingly, @lush2022 did test for a relationship between PC and illusion sensitivity using the Müller-Lyer illusion (in which the arrangement of the arrowheads flanking two lines makes them appear as having different lengths), and reported evidence in favour of an absence of correlation between the two measures. This finding was interpreted as indicative of the cognitive impenetrability of illusions, implying that the effect is driven by low-level processes and therefore not influenced by top-down mechanisms such as PC. The goal of this study is thus to replicate the results from @lush2022 pointing to an absence of a relationship between phenomenological control and illusion sensitivity, by generalising them to a different illusion task that would encompass other illusion types (see @tbl-table1).**Our paradigm extends that of Lush’s by presenting this task as a gamified paradigm hopefully motivating participants to perform at their best and to remain engaged throughout**.

@@

The experiment’s setup follows **a modified version** of the born-open principle (De Leeuw, 2023), **with adjustments made to ensure data privacy, confidentiality and transparency**. The online experiment, implemented entirely using JsPsych (De Leeuw, 2015), has its code stored on GitHub and will leverage the power of the platform to host the experiment for free. **Participant’s raw data files will be sent directly to a private OSF repository. These files will then undergo pre-processing to anonymize the data by removing any sensitive information, such as Prolific and SONA IDs. The anonymized data will then be saved on GitHub, where it will be accessible for analysis, along with the pre-processing script**.

Participants will be presented with a consent form followed by demographic questions (gender, education level, age, and ethnicity). **Although these variables are not analyzed in the current study, they will be made available to provide a detailed and thorough description of the sample and to ensure consistency with previous studies using the paradigm (Makowski et al., 2023), as well as maximizing reusability**. Participants will then be administered the PCS and the Illusion Game task (IG) in a counterbalanced order.

@@

The task is an adaptation of the one used in @makowski2023 to make it shorter and more reliable, in which participants must make perceptual judgments (e.g., "which red line is the longer") as quickly and accurately as possible. It includes 3 illusion types, namely Ebbinghaus, Müller-Lyer, and Vertical-Horizontal. The procedure encompasses 2 sets of 80 trials for each illusion type. Each set will include, in a random order, the 3 blocks of illusion types, in which trials are separated by a fixation cross, temporally (uniformly sampled duration of 500 - 1000s) and spatially jittered (around the centre of the screen in a radius of a 1 cm) to attenuate its potential usefulness as a reference point. After each illusion type block, a score is presented (computed as a scaled Inverse Efficiency Score) as a gamification mechanism to increase motivation to perform to the best of one's abilities. **To mitigate the speed/accuracy trade-off, inherent to most decision-making tasks, we will maintain a double-constrained instruction by emphasizing with equal weight to “be as fast as possible without making errors”**.

@@

The two sets of 3 illusion blocks will be separated by 2 short questionnaires acting as a break, namely the IPIP-6 [@sibley2011mini], measuring 6 personality traits with 24 analogue scales items, and the PID-5 [@krueger2011], measuring 5 maladaptive personality traits with 25 Likert scales items.**These questionnaires are strategically included to provide a break between the two cognitively taxing blocks of comparable activity and duration and to maximize the reusability of the data by maintaining consistency with paradigms used previously (Makowski et al., 2023)**.
